### 2014-03-27 paper.tex
A more detailed plan of the paper, plus some text
on pull arrays. I've also added a rosetta stone to help
translate between MoA and Pull arrays

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{longtable}
\usepackage{stmaryrd}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Title Text}
\subtitle{Subtitle Text, if any}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

\section{Overview of MoA}

\section{Overview of Pull arrays}

The functional programming community has recently gained an increased
interest in high performance parallel array programming
\cite{keller2010regular,Axelsson:2010:Feldspar,Mainland:2010:Nikola,Svensson:2011:Obsidian,Claessen:2012:Expressive,Ankner:2013:AnEDSL,lippmeier2011efficient}. One
of the central abstractions in this line of work is the \emph{pull
  array}, also know as delayed array. The can be defined as follows
(we use Haskell \cite{marlow2010haskell} to demonstrate functions
programs).

\begin{verbatim}
data Pull a = Pull { ixf    :: (Int -> a)
                   , length :: Int
                   }
\end{verbatim}

Arrays are represented as functions from index to element. We refer to
this function as an \emph{index function}. Additionally arrays also
has a length.

This representation has several advantages:
\begin{itemize}
\item \emph{Parallelism}. Since each element is computed
  independently, they can easily be computed in parallel.
\item \emph{Fusion}. When functions on pull arrays are composed, the intermediate
  arrays can be automatically removed. 
\item \emph{Compositionality}. Pull arrays provide a wealth of
  compositional and highly reusable combinators. These can be composed
  together to write high-level programs which typically are easier to
  understand than monolithic array processing code.
\end{itemize}

\begin{figure}
\begin{verbatim}
map :: (a -> b) -> Pull a -> Pull b
map f (Pull ixf l) = Pull (f . ixf) l
\end{verbatim}
\caption{Example functions on pull arrays}
\end{figure}

\subsection{Fusion}

\subsection{Higher dimensions}

\section{Comparison}

\subsection{Types}

\subsection{\(\psi\) reductions and fusion}

\subsection{Higher dimensions}

\subsection{Functions}

\begin{tabular}{|@{}l|l@{}|}
\hline
MoA & Feldspar
\\
\hline
δ & dim . extent
\\
ρ & extent
\\
ιξⁿ & -- generalized enumFromTo
\\
ψ & index
\\
rav & -- flattens to a one-dimensional array
\\
γ & toIndex
\\
γ' & fromIndex
\\
$s \: \hat{ρ} \: ξ$ & reshape
\\
π x & product
\\
τ & size
\\
++ & ++
\\
ξ₁ f ξ₂ & zipWith f ξ₁ ξ₁
\\
σ f ξ & map f ξ
\\
ξ f σ & map f ξ
\\
$\bigtriangleup$ & take
\\
$\bigtriangledown$ & drop
\\
$op^{red}$ & fold
\\
Φ & reverse
\\
Θ & rotate
\\
$\varobslash$ & -- generalized transpose
\\
$f Ω_d ξ$ & -- Apply $f$ along the dimension $d$ of $ξ$
\\
\hline
\end{tabular}

\subsection{Example programs}

\section{Related work}

\section{Future work}

\section{Conclusion}

\acks

Acknowledgments.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{bibliography}

% The bibliography should be embedded for final submission.

%% \begin{thebibliography}{}
%% \softraggedright

%% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%% P. Q. Smith, and X. Y. Jones. ...reference text...

%% \end{thebibliography}

\end{document}

### 2014-03-31 intromoa.tex
I have added psi reduction example.
If ok with you let's use the same example. As mentioned in email if we
introduce omega we'll lose everyone during psi reduction.

% See the ``Article customise'' template for come common customisations
\newcommand{\cat}{+\!\!\!+}
%\newcommand{\cat}{\mbox{$+\!\!+_s$}}
\newcommand{\Take}{\mbox{Take}}
\newcommand{\take}{\,\bigtriangleup\,}
\newcommand{\Drop}{\mbox{Drop}}
\newcommand{\drop}{\,\bigtriangledown\,}
\newcommand{\rshp}{\,\widehat{\Rho}\;}
\newcommand{\Rho}{\,\rho\,}
\newcommand{\Tau}{\,\tau\,}
\newcommand{\Dim}{\,\delta\,}
\newcommand{\transpose}{\bigcirc\!\!\!\!\!\backslash\;}
\newcommand{\kron}{\bigcirc\,\!\!\!\!\!\!\times\;}
\newcommand{\Ravel}{\,\mbox{\tt rav}\,}
\newcommand{\Gradeup}{\,\mbox{\tt gu}\,}
%\newcommand{\cat}{+\!\!\!\!+}

\newtheorem{definition}{Definition}
%\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{theorem}{Theorem}

\noindent

MoA, an acoronym for {\em A Mathematics of Arrays\cite{thesis}}, is a theory of arrays with both an algebra and a calculus of 
indexing based on shapes, i.e. sizes of each dimension. The index calculus, refered to as {\em The Psi Calculus}, 
gets its names from the {\em Psi}  function which is the foundational function in the theory. The name {\em Psi} 
was inspired by Dirac's {\em Psi Vectors} and his communications with Einstein to have a mathematical way to reason
about them. These results began with the awareness that Iverson's APL\cite{iverson62}
was inspired by Sylvester's Universal Algebra\cite{sylvester} and the fact that he and Cayley invented matrices.
The programming language, APL, embodied Ken's notation into an interpretive computing environment. It was at this point
that theoretical anamolies were introduced. More on this later.

Not only did  Ken believe that APL's array algebra was the ideal basis for a programming language, he (and others) believed that
an ideal machine language supported array operations. This led to HP building an 
APL machine based on Phil Abram's dissertation\cite{abrams70}. Abrams was the first to identify ways 
to define an algebra of arrays based on shapes; others followed\cite{lyon,guibas}. He believed that if array operations were based on 
indexing with  shapes, all  indexing in the array expression could be  composed
and  thus could minimize the creation (and deletion) of temporary arrays,
a huge overhead. One of the reasons HP's APL Machine did not succeed was that this analysis was done at run time.
It was  Alan Perlis  who realized that these optimizations should be done at compile time\cite{perlis}, others
followed\cite{miller,budd84}. Perlis was also the first to recognize that there were too many anamolies in both APL's
algebra and environment; he recommended a functional array calculator based on the lambda 
calculus\cite{tu2,tu3}. This awareness led to Mullin's collaborations with Klaus Berkling to combine MoA, Psi Calculus, and
the Lambda Lalculus, theoretically an ideal enviornment\cite{berkling} and basis for a 
functional programming language with 
arrays\cite{mul91,budd91,hm92,mul93,mul94,mul96}.

When compiler support was implemented in APL,
the main limitation was that the anomalies in the language~\cite{gerhart} 
made it difficult to optimize and impossible to verify correctness of 
programs.

MoA and the Psi Calculus, as a theory  has  also been used to describe and verify 
hardware\cite{hardy1,hardy2,fftharry}, describe the partitioning and
scheduling of array operations (and their costs)\cite{mdst93,lee95,aachen96,coffin94,ll2,mcmahon95,haleh98}, 
and the formulation of parallel and distributed processing\cite{Mul03}. 
MoA and $\psi$-Calculus, can be used to abstract complex processor memory layouts 
for tensor/array expressions, reduce these expressions first to a semantic 
normal form (Denotational Normal Form-DNF) then to an operational normal form 
(ONF). The ONF describes how to {\em build} the code using 
{\em start, stop, stride}, and {\em count}: a universal machine abstraction.
This is possible by {\em abstractly} lifting the dimension of an expression to include all aspects
of a processor, memory, communication topology. . 
%
%
%
%\bibliographystyle{plain}
%\bibliography{paper1,hpf,workshop}
\section*{Elements of the theory}
\subsection*{\label{indshp} Indexing and Shapes}

The central operation of MoA is the indexing function
\[
                    p \psi A \]
in which a vector of $n$ integers $p$ is used to select an item of 
the $n$-dimensional array $A$.
The operation is generalized to select a partition of $A$, 
so that if $q$ has only $k < n $ components
then
\[                    q \psi A \]
is an array of dimensionality $n-k$ and $q$ selects among the possible 
choices for the first $k$ axes. In MoA zero origin indexing is assumed.
For example, if $A$ is the $3$ by $5$ by $4$ array
\[ \left [
\begin{array}{lrrrr}
0 & 1 &  2 & 3  \\
4 & 5 & 6 & 7 \\
8 & 9 & 10 & 11  \\
12 & 13 & 14 & 15 \\
16 & 17 & 18 & 19 
\end{array} \right ] 
\left [ \begin{array}{rrrr}
20 & 21 & 22 & 23 \\
24 & 25 & 26 & 27 \\
28 & 29 & 30 & 31 \\
32 & 33 & 34 & 35 \\
36 & 37 & 38 & 39 
\end{array} \right ]
\left [ \begin{array}{rrrr}
40 & 41 & 42 & 43 \\
44 & 45 & 46 & 47 \\
48 & 49 & 50 & 51 \\
52 & 53 & 54 & 55 \\
56 & 57 & 58 & 59
\end{array} \right ] \]
then
\[
     <1> \psi A = \left [  \begin{array}{rrrr}
                   20 & 21 & 22 & 23 \\
                   24 & 25 & 26 & 27 \\
                   28 & 29 & 30 & 31 \\
                   32 & 33 & 34 & 35 \\
                   36 & 37 & 38 & 39 
                    \end{array} \right ] \]

\[     <2 \; 1> \psi A \; \; = \; \; < \; 44 \;\; 45 \;\; 46 \;\; 47 \; > \]

\[      <2\; 1\; 3> \psi A \; \; = \; \; 47 \]

Most of the common array manipulation operations found in languages like Fortran90, Matlab, ZPL, etc.,
 can be defined from $\psi$ and a few elementary vector operations.

	We now introduce notation to permit us to define $\psi$ formally and to develop the
{\em Psi Correspondence Theorem}~\cite{muljen94}, which is central to the effective exploitation of MoA in array
computations. We will use $A, B, ...$ to denote an array of numbers (integer, real, or complex).
An array's dimensionality will be denoted by $d_A$ and will be assumed to 
be $n$ if not specified.

	The shape of an array $A$, denoted by $s_A$, is a vector of integers of length $d_A$,
each item giving the length of the corresponding axis. The total number of items in an
array, denoted by $t_A$, is equal to the product of the items of the shape. The subscripts
will be omitted in contexts where the meaning is obvious.

A full index is a vector of $n$ integers that describes one position 
in an $n$-dimensional
array. Each item of a full index for $A$ is less than the corresponding item of
$s_A$. There are precisely $t_A$ indices for an array(due to a
zero index origin). A partial index of $A$ is a vector of $0 \leq  k < n $
integers with each item less than the corresponding item of $s_A$.

	We will use a tuple notation (omitting commas) to describe vectors of a fixed length.
For example,
\[                  <i \; j \; k> \]
denotes a vector of length three. $<>$ will denote the empty vector which is 
also sometimes written as $\Theta$.

	For every $n$-dimensional array $A$, there is a vector of the items of $A$,
which we denote by the corresponding lower case letter, here $a$. The length of the vector of
items is $t_A$. A vector is itself a one-dimensional array, whose shape is the one-item
vector holding the length. Thus, for $a$, the vector of items of $A$, the
shape of $a$ is
\[                     s_a =  \; < t_A> \]
and the number of items or total number of components
$a$\footnote{We also use $\tau a , \delta a,$ and $\rho a$ to denote
total number of components, dimensionality and shape of a. } is
\[                    t_a =  t_A . \]

	The precise mapping of $A$ to $a$ is determined by a one-to-one 
ordering function, $gamma$.  Although the choice of ordering is arbitrary, it is 
essential in the following that a specific one be assumed. By convention 
we assume the items of $A$ are placed in $a$ according to the lexicographic 
ordering of the indices of $A$. This is often referred to 
as $row \; major \; ordering$. 
Many programming languages lay out the items of multidimensional
arrays in memory in a contiguous segment using this ordering. 
Fortran uses the ordering corresponding to a transposed array in which the 
axes are reversed {\em column major}. Scalars are introduced as arrays with an empty shape vector.

There are two equivalent ways of describing an array $A$:
\begin{description}
\item[(1)] by its shape and the vector of items, i.e.  $A = \{s_A , a\}$, or
\item[(2)] by its shape and a function that defines the value at every index $p$.
\end{description}
These two forms have been shown to be formally equivalent~\cite{jenk94}.
We wish to use the second form in defining functions on multidimensional
arrays using their Cartesian coordinates (indices). 
The first form is used in describing
address manipulations to achieve effective computation.

To complete our notational conventions, we assume that $p,q,...$ will 
be used to denote
indices or partial indices and that $u,v,..$ will be used to denote 
arbitrary vectors of integers.
In order to describe the $i_{th}$ item of a vector $a$, either $a_i$ or 
$a[i]$ will be used. If $u$ is a
vector of $k$ integers all less than $t_A$, then $a[u]$
will denote the vector of length $k$, whose items are the items of $a$ at 
positions $u_j$, $j = 0,...,k-1.$

Before presenting the formal definition of the $\psi$ indexing function we 
define a few functions on vectors:
\begin{tabbing}

\hspace{2cm}\= $u \;\cat\; v$ \hspace{1cm} \=		\=catentation of vectors $u$ and $v$ \\
\>$u$ + $v$\>		\>itemwise vector addition assuming $t_u = t_v$ \\
\>$u \times v$ \>	\>	itemwise vector multiplication \\
\>$n$ + $u$, $u$ + $n$ \>	addition of a scalar to each item of a vector \\
\>$n \times u$, $u \times n$ \>   multiplication of each item of a vector by a scalar \\
\>$\iota \; n$ \>\>  the vector of the first n integers starting from $0$ \\
%\footnote{$\iota \vec n$ is also defined and  produces an array of 
%cartesian coordinates~\cite{mul88}.}\\
\>$\pi \; v$ \> \>a scalar which is the product of the components of v \\
\>$k \;\take \;u$ \>\>	when $k \geq 0$ the vector of the first $k$ items of $u$,(called {\em take}) \\
\> \> \>and when $k<0$ the vector of the last $k$ items of $u$ \\
\>$k \;\drop \;u$ \>  \> when $k \geq 0$ the vector of $t_u - k$ last items of $u$,(called{ \em drop}) \\
\> \> \>and when $k<0$ the vector of the first $t_u - |k|$ items of $u$ \\
\>$k \;\theta \;u$ \>  \>  when $k \geq 0$  the vector of $(k  \drop  u) \cat ( k \take u ) $\\
\> \> \>and when $k<0$ the vector or $(k \take u ) \cat ( k \drop u ) $
\end{tabbing}

\begin{definition}
Let A be an n-dimensional array and p a vector of integers.
If p is an index of A, 
\[             p \psi A = a[\gamma(s_A,p)], \]
   where 
\begin{eqnarray}
      \gamma(s_A,p) & = & x_{n-1}  \;\;\;\;\; \mbox{defined by 
the recurrence}    \nonumber \\ 
                 x_0 & = & p_0, \nonumber \\ 
              x_j & = & x_{j-1} * s_j + p_j, \; \; \; j=1,...,n-1. \nonumber
\end{eqnarray}

If $p$ is partial index of length $k < n,$
\[             p \psi A = B \]
where the shape of $B$ is
\[         s_B = k \drop s_A,    \] 
and for every index $q$ of $B$,
\[         q \psi B = (p \cat q) \psi A \]
\end{definition}
The definition uses the second form of specifying an array to define the result
of a partial index. For the index case, the function $\gamma(s,p)$ 
is used to convert an index $p$ to an
integer giving the location of the corresponding item 
in the row major order list of items of an array of shape $s$. The
recurrence computation for $\gamma$ is the one used in most compilers 
for converting an
index to a memory address~\cite{djr}.

\begin{corollary}
$<> \psi$ A = A.
\end{corollary}

The following theorem shows that a $\psi$ selection with a partial index can
be expressed as a composition of $\psi$ selections.

\begin{theorem}
Let A be an n-dimensional array and p a partial index so 
that $p = q \cat r.$ Then
\[                 p \psi A = r \psi (q \psi A) . \]

Proof: The proof is a consequence of the fact that for vectors u,v,w
\[           (u \cat v) \cat w = u \cat (v \cat w). \]
If we extend p to a full index by $p \cat p',$ then

\begin{eqnarray}
             p' \psi (p \psi A) & = & (p \cat p') \psi A \nonumber \\
                             & = & ((q \cat r) \cat p') \psi A \nonumber \\
                             &  =  & (q \cat (r \cat p')) \psi A \nonumber \\
                            &  =  & (r \cat p') \psi (q \psi A) \nonumber \\
                            &   = & p' \psi ( r \psi (q \psi A)) \nonumber \\
                          p \psi A & = & r \psi (q \psi A) \nonumber                   
\end{eqnarray}

\end{theorem}
{\em which completes the proof.}

We can now use {\em psi} to define other operations on arrays. 
For example, consider definitions of {\em take} and {\em drop} for multidimensional arrays.
\begin{definition}[take: $\take$]
Let A be an n-dimensional array, and k a non-negative integer
such that $0 \leq k < s_0$. Then
\[            k \take A = B \]
where
\[            s_B = <k> \cat (1 \drop s_A) \]
and for every index p of B,
\[            p \psi B = p \psi A . \]

\end{definition}
(In MoA $\take$ is also defined for negative integers and is generalized to any vector u with its
absolute value vector a partial index of A. The details are omitted here.)
\begin{definition}[reverse: $\Phi$]
 Let A be an n-dimensional array. Then
\[             s_{\Phi A} = s_A \]
and for every integer i, $0 \leq i < s_0, $
\[             <i> \psi \Phi A = <s_0 - i - 1> \psi A. \]
\end{definition}
This definition of $\Phi$ does a reversal of the 0th axis of A.

Note
also that all operations are over the 0th axis. The operator $\Omega$~\cite{mul88} extends operations over all other dimensions.
\subsubsection*{Example}
Consider the evaluation of the following expression using the
3 by 5 by 4 array, $A$, introduced in Section~\ref{indshp}.
\begin{eqnarray}
             <1\;\; 2> \psi (2 \take \Phi A) 
\end{eqnarray}
where A is the array given in the previous section.
The shape of the result is 
\begin{eqnarray}
                & &  2 \drop s_{(2 \take \Phi A)} \nonumber \\
                & = & 2 \drop (<2> \cat (1 \drop s_{\Phi A} )) \nonumber \\
               & = & 2 \drop (<2> \cat (1 \drop s_A)) \nonumber \\
               & = & 2 \drop (<2> \cat <5\; 4>) \nonumber \\
               & = & 2 \drop <2\; 5 \;4> \nonumber \\
              &  = &<4> . 
\end{eqnarray}
The expression can be simplified using the definitions:
\begin{eqnarray}
          &&   <1\; 2> \psi (2 \take \Phi A) \nonumber \\
           & = &  <1 \; 2> \psi \Phi A \nonumber \\
         &   = & <2> \psi (<1> \psi \Phi A) \nonumber \\
         &  = & <2> \psi (<3 - 1 - 1 > \psi A) \nonumber \\
        &   = & <1 \; 2> \psi A  \nonumber \\
\end{eqnarray}
This process of simplifying the expression for the item in terms of
its Cartesian coordinates is called {\em Psi Reduction}. The operations of MoA
have been designed so that all expressions can be reduced to a minimal
normal form~\cite{mul88}.
Some MoA operations defined by {\em psi} are found in Fig.\ref{moa}.

%\break
\begin{figure}[h]
\scriptsize
\begin{tabular}{|l|l|l|}
\hline\hline
{\bf Symbol} & {\bf Name}& {\bf Description} \\ \hline
$\delta$ & Dimensionality & Returns the number of dimensions of an array.\\ \hline
$\rho$ & Shape & Returns a vector of the upper bounds \\
& &or sizes of each dimension in an array. \\ \hline
$\iota \xi^n$  & Iota & When $n=0$(scalar),
returns a vector containing elements\\
& & $0$, to $\xi^0 - 1$. When $n = 1$(vector), returns an \\
& & array of indices  defined by the shape vector $\xi^1$ \\ \hline
$\psi$ & Psi & The main indexing function of the Psi Calculus \\
& & which defines all operations in MoA.   Returns a scalar \\
& & if a full index is provided, a sub-array otherwise. \\ \hline
\mbox{rav} & Ravel & vectorizes a multi-dimensional array based   \\
&&on an  array's layout($\gamma_{row},\gamma_{col},\gamma_{sparse}, ...$)\\ \hline
$\gamma$ & Gamma & Translates indices into offsets given a shape. \\ \hline
$\gamma^{'}$ & Gamma Inverse & Translates offsets into indices given a shape. \\ \hline
$\vec s \rshp \xi $ & Reshape& Changes the shape vector of an array,
 possibly affecting  \\
& &its dimensionality. Reshape depends on
layout($\gamma$). \\ \hline
$\pi \vec x$ & Pi & Returns a scalar and is equivalent to $\prod_{i=0}^{(\tau x)-1}x[i]$ \\ \hline
$\tau$& Tau & Returns the number of components in an array,($\tau \xi \equiv
\pi (\rho \xi) $) \\ \hline
$\xi_l \cat \xi_r $ & Catenate& Concatenates two arrays over their
primary axis. \\ \hline
$\xi_l f \xi_r $ & Point-wise  & A data parallel application of $f$ is performed \\
&Extension &  between all elements of the arrays.\\ \hline
$\sigma f \xi_r $ & Scalar Extension & $\sigma$ is used with
every component of $\xi_r$ in the data parallel \\
$\xi_l f \sigma$ & &application of $f$. \\ \hline
$\take$&Take& Returns a sub-array from the beginning or end of an array\\
& & based on its argument being positive or negative. \\ \hline
$\drop$ & Drop & The inverse of Take \\ \hline
$_{op} \mbox{red}$&Reduce& Reduce an array's dimension by one by applying \\
& & op over the primary axis of an array. \\ \hline
$\Phi$& Reverse& Reverses the components of an array. \\ \hline
$\Theta $& Rotate & Rotates, or shifts cyclically, components of an
array. \\ \hline
$\transpose$ & Transpose & Transposes the elements of an array based on \\
&&a given permutation vector \\ \hline
$\Omega$& Omega& Applies a unary or binary function to array argument(s) \\
&&given partitioning information. $\Omega$ is used to perform all operations \\&& (defined over the primary axis only) over all dimensions. \\ \hline
\end{tabular}
\caption{\label{moa}Summary of MoA Operations}
\end{figure}

\section*{Higher Order Operations}

Thus far operation on arrays, such as concatenation, rotation, etc., have been 
performed over their 0th dimensions.  We introduce the higher order binary 
operation $\Omega$, which is defined when its left argument is a unary or binary
operation and its right argument is a vector describing the dimension upon 
which operations are to be performed, or which sub-arrays are used in 
operations.  The dimension upon which operations are to be performed is often 
called the {\it axis} of operation.  The result of $\Omega$ is a unary or 
binary operation.

### 2014-04-01 intromoa.tex
more edits
latest version, still many bib files. Awaiting your review.

@@
 The programming language, APL, embodied Ken's notation into an interpretive computing environment. It was at this point
-that theoretical anamolies were introduced. More on this later.
+that theoretical anamolies were introduced.
 
@@
 APL machine based on Phil Abram's dissertation\cite{abrams70}. Abrams was the first to identify ways 
-to define an algebra of arrays based on shapes; others followed\cite{lyon,guibas}. He believed that if array operations were based on 
+to define an algebra of arrays based on shapes; others attempted to
+put closure on his work\cite{lyon,guibas}. Abrams believed that if array operations were based on 
 indexing with  shapes, all  indexing in the array expression could be  composed
@@
 a huge overhead. One of the reasons HP's APL Machine did not succeed was that this analysis was done at run time.
-It was  Alan Perlis  who realized that these optimizations should be done at compile time\cite{perlis}, others
-followed\cite{miller,budd84}. Perlis was also the first to recognize that there were too many anamolies in both APL's
-algebra and environment; he recommended a functional array calculator based on the lambda 
+It was  Alan Perlis, and later others\cite{miller,budd84},  who realized that these optimizations should be done at compile time.
+Attempts at compilation led Perlis to realize that  anomalies in the language~\cite{gerhart}
+made it difficult to optimize and impossible to verify correctness of 
+programs leading him to recommend a functional array calculator based on the lambda 
 calculus\cite{tu2,tu3}. This awareness led to Mullin's collaborations with Klaus Berkling to combine MoA, Psi Calculus, and
@@
 
-When compiler support was implemented in APL,
-the main limitation was that the anomalies in the language~\cite{gerhart} 
-made it difficult to optimize and impossible to verify correctness of 
-programs.
-
-MoA and the Psi Calculus, as a theory  has  also been used to describe and verify 
+MoA and the Psi Calculus, as a theory,  have  also been used to describe and verify 
 hardware\cite{hardy1,hardy2,fftharry}, describe the partitioning and
@@
 for tensor/array expressions, reduce these expressions first to a semantic 
-normal form (Denotational Normal Form-DNF) then to an operational normal form 
+normal form (Denotational Normal Form-DNF) then to an Operational Normal Form 
 (ONF). The ONF describes how to {\em build} the code using 
@@
 This is possible by {\em abstractly} lifting the dimension of an expression to include all aspects
-of a processor, memory, communication topology. . 
+of a processor, memory, communication topology. For example, to map a 2-d array expression to
+n  processors we would lift the dimenion to 3-d. 
 %
@@
 %\bibliography{paper1,hpf,workshop}
-\section*{Elements of the theory}
+\subsection*{Elements of the theory}
 \subsection*{\label{indshp} Indexing and Shapes}
@@
 
-\hspace{2cm}\= $u \;\cat\; v$ \hspace{1cm} \=		\=catentation of vectors $u$ and $v$ \\
+\hspace{.25cm}\= $u \;\cat\; v$ \hspace{1cm} \=		\=catentation of vectors $u$ and $v$ \\
 \>$u$ + $v$\>		\>itemwise vector addition assuming $t_u = t_v$ \\
@@
 
-\section*{Higher Order Operations}
+\subsection*{Higher Order Operations}
 

### 2014-04-01 paper.tex
I have added psi reduction example.
If ok with you let's use the same example. As mentioned in email if we
introduce omega we'll lose everyone during psi reduction.
paper.tex changes
Affiliation, etc
Affiliation
Section about fusion for Pull arrays

@@
            {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
+\authorinfo{Lenore M. Mullin}
+           {Emeritus Professor\\Department of Computer Science \\
+University at Albany, SUNY}
+           {lenore@cs.albany.edu}
 
@@
 \section{Overview of MoA}
+\input{intromoa}
 
@@
 \subsection{Fusion}
+
+As mentioned, one of the most important advantages of Pull arrays is
+that they support
+fusion\cite{gill1993short,axelsson2010feldspar,keller2010regular}. Fusion
+is a program transformation which removes intermediate data structures
+and is a decendant of deforestation
+\cite{wadler1990deforestation}. Fusion supercedes transformations such
+as loop fusion in optimizing compilers for imperative languages.  Pull
+arrays make it particularly simple to achieve fusion. It simply
+amounts to inlining the definitions of functions.
+
+As an example of fusion consider the example
+\verb!map (*2) (map (*5) arr)!, where each element of the array
+\verb!arr! is first multiplied by five and then again by
+two. Conceptually, there is an intermediate array created after the
+first \verb!map! but thanks to fusion it will be removed. Fusion will proceed
+as follows:
+
+\begin{verbatim}
+map (*2) (map (*5) arr)
+=> { arr }
+map (*2) (map (*5) (Pull ixf l)
+=> { map }
+map (*2) (Pull ((*5) . ixf) l)
+=> { map }
+Pull ((*2) . (*5) . ixf) l)
+\end{verbatim}
+
+The result is a single Pull array, the intermediate array has been removed.
+
+The correctness of fusion relies crucially on the language being
+\emph{pure}. If side effects where allowed in the indexing function
+fusion would not be correct.
+
+There are two different ways of implementing Pull arrays in the
+literature. The simplest one comes from using Pull arrays in an
+embedded langauge, i.e. when the array language is implemented as a
+library in another host language. Examples of such languages are
+Pan\cite{elliott2003compiling}, Feldspar\cite{Axelsson:2010:Feldspar},
+Obsidian\cite{Svensson:2011:Obsidian} and
+Nikola\cite{Mainland:2010:Nikola}. By using the technique of
+implementing Pull arrays as \emph{shallow embeddings} on top of a
+\emph{deeply embedded} code language, fusion comes completely for free
+\cite{svenningsson2013combining}.
+
+In the array library Repa\cite{keller2010regular}, compiler supported
+rewrite rules are used to implement fusion \cite{jones2001playing}.
 
@@
 \bibliographystyle{abbrvnat}
-\bibliography{bibliography}
+\bibliography{bibliography,hpf,paper1,workshop}
 

### 2014-04-01 paper.tex
More writing on the Pull intro

@@
 
-\authorinfo{Name1}
-           {Affiliation1}
-           {Email1}
+\authorinfo{Josef Svenningsson}
+           {Department of Computer Science of Engineering\\Chalmers University of Technology}
+           {josef.svenningsson@chalmers.se}
 \authorinfo{Lenore M. Mullin}
@@
 \begin{verbatim}
+-- Mapping a function over an array
 map :: (a -> b) -> Pull a -> Pull b
 map f (Pull ixf l) = Pull (f . ixf) l
+
+-- Apply a binary operation itemwise on two arrays
+zipWith :: (a -> b -> c)
+        -> Pull a -> Pull b -> Pull c
+zipWith f (Pull ixf1 l1) (Pull ixf2 l2)
+  = Pull (\i -> f (ixf1 i) (ixf2 i)) (min l1 l2)
+
+-- Takes the n first elements of an array
+take :: Int -> Pull a -> Pull a
+take n (Pull ixf l) = Pull ixf (min n l)
+
+-- Drops the n first elements of an array
+drop :: Int -> Pull a -> Pull a
+drop n (Pull ixf l)
+  = Pull (\i -> ixf (i + n)) (max (l - n) 0)
+
+-- Rotate an array k steps to the left
+rotate :: Int -> Pull a -> Pull a
+rotate k (Pull ixf l)
+  = Pull (\i -> ixf ((i + k) `mod` l)) l
 \end{verbatim}
@@
 \subsection{Higher dimensions}
+
+The first uses of Pull arrays used a fixed number of dimensions,
+typically either one or two. However, the Repa library introduced a
+new expressive notion of higher dimensionality for Push arrays
+\cite{keller2010regular}. The number dimensions of an array, referred
+to as the \emph{shape}, are determined statically in the type
+system. While this might seem very constraining, a key innovation in
+Repa is the notion on \emph{shape polymorphism} which makes it
+possible to write functions which accepts arrays of any shape. Below
+shows the definition of the type of shapes. It is a variation on how
+it is represented in Feldspar.
+
+\begin{verbatim}
+data Z
+data i :. sh
+
+data Shape sh where
+  Z :: Shape Z
+  (:.) :: Shape tail -> Int -> Shape (tail :. Int)
+\end{verbatim}
+
+The first two data declarations introduces new types, \verb!Z! and
+\verb!:.!, without any constructors. They are used to create
+snoc-lists on the type level, the definition of \verb!Shape!.  The
+type \verb!Shape! is a recursive data type with two constructors. The
+first constructor \verb!Z! indicates that the shape has zero
+dimensions.  The second constructor \verb!:.! adds one dimension to
+another shape, and stores how many elements there are in that
+dimension. The type \verb!Shape! has a type parameter which is a snoc
+list with one element for each dimension.
+
+\begin{verbatim}
+data Pull sh a = Pull (Shape sh -> a) (Shape sh)
+
+map :: (a -> b) -> Pull sh a -> Pull sh a
+map f (Pull ixf sh) = Pull (f . ixf) sh
+
+zipWith :: (a -> b -> b) -> 
+           Pull sh a -> Pull sh b -> Pull sh c
+zipWith f (Pull ixf1 sh1) (Pull ixf2 sh2)
+  = Pull (\sh -> f (ixf1 sh) (ixf2 sh2))
+\end{verbatim}
 

### 2014-04-04 paper.tex
A first stab at an introduction.

@@
 \authorinfo{Lenore M. Mullin}
-           {Emeritus Professor\\Department of Computer Science \\
+           {Emeritus Professor\\Department of Computer Science\\
 University at Albany, SUNY}
@@
 
+Computing with arrays has been a constantly important paradigm in the
+history of computing. For this reason we have seen a plethora of
+formalisms and languages being developed to aid theoriticians and
+developers to produce more suitable abstractions and faster programs.
+Examples include Fortran, APL and Matlab but the list goes on and on.
+Having many different options to choose from can be a good thing but
+it can also lead to a balkanization of the community. When researcher
+dedicated to different formalisms can no longer talk to each other it
+inhibits collaborations and limits the impact of innovatation.
+
+In this paper we aim to rectify some of the fragmentation of the
+high-performance array programming community by relating two
+particular formalisms: Mathematics of Arrays and Pull arrays. On the
+surface they appear rather different but at their core they share many
+principles and properties which we will examine in this paper. In
+particular we will highlight the following points:
+\begin{itemize}
+\item We give an introduction to both MoA (section \ref{sec:moa}) and Pull
+  arrays (section \ref{sec:pull}).
+\item The two formalisms are compared with respect to types (section
+  \ref{sec:types}), fusion/\(\psi\)-reduction (section
+  \ref{sec:normalization}), how they deal with higher dimensions
+  (section \ref{sec:highdim}) and notation (section \ref{sec:notation}).
+\end{itemize}
+
+We would like to stress that we are not looking to declare a
+winner. Instead our intention is that this paper can serve as a help
+for the two communities to be able to learn about each others work.
+
 \section{Overview of MoA}
+\label{sec:moa}
+
 \input{intromoa}
@@
 \section{Overview of Pull arrays}
+\label{sec:pull}
 
@@
 to as the \emph{shape}, are determined statically in the type
-system. While this might seem very constraining, a key innovation in
+system. The advantage of having the shape as part of the type is that
+any program which produces arrays with ill-defined shapes can be ruled
+out. A potential downside is that enforcing shapes in the type system might
+be overly constraining and exclude useful programs. A key innovation in
 Repa is the notion on \emph{shape polymorphism} which makes it
@@
 \subsection{Types}
+\label{sec:types}
 
-\subsection{\(\psi\) reductions and fusion}
+\subsection{\(\psi\)-reductions and fusion}
+\label{sec:normalization}
 
 \subsection{Higher dimensions}
+\label{sec:highdim}
 
 \subsection{Functions}
+\label{sec:notation}
 

### 2014-04-04 intromoa.tex
changes for psi reduction

@@
 For example, if $A$ is the $3$ by $5$ by $4$ array
+\small
 \[ \left [
@@
 \end{array} \right ] \]
+\normalsize
 then
@@
 \begin{figure}[h]
-\scriptsize
+%\scriptsize
+\tiny
 \begin{tabular}{|l|l|l|}
@@
 binary operation.
+
+\subsection*{From Semantic to Operational Forms}
+
+%Section with Psi  Correspondence Theorem and thinking in arrays *\
+
+Consider the expression
+
+\begin{eqnarray}
+X & = & (2 \take ( \Phi A)) \times ( 1 \drop (\Phi A)) 
+\end{eqnarray}
+That is, take the first two planes after reversing A along the primary axis then multiply that array by the last two 
+planes after reversing A. We want to concurrently perform both 5 by 4 multiplications. 
+A is the array previously defined with shape $<3\;5\;4>$.
+Suppose also that A is in shared memory and accessible to 2 processors.
+First, we {\em Psi-Reduce} our expression to its Denotational Normal Form (DNF) or semantic 
+normal form. This normal form can be used to prove the equivalence of  two array expressions.
+
+Get shape: 
+\begin{eqnarray}
+\rho X & = & \rho (2 \take (\Phi A)) \nonumber \\
+& = & 2 \cat ((1\drop (\rho ( \Phi A)))) \nonumber \\
+&=& 2 \cat ((1 \drop (\rho  A))) \nonumber \\
+&=& 2 \cat <\;5\;4> \nonumber \\
+&=& <2\;4\;5>
+\end{eqnarray}
+Now that we have the {\em shape}, we can {Psi-Reduce}.
+\[ \forall \;\;\; i,j,k \;\;\; \ni \;\;\; 0 \;\;\;\leq i,j,k < \;\;\;<2\;4\;5> \]
+\[<i\;j\;k> \psi X \]
+\begin{eqnarray}
+& =&<i\;j\;k> \psi (2 \take ( \Phi A)) \times ( 1 \drop (\Phi A)) \nonumber \\
+&=&<j\;k> \psi (<i> \psi (2 \take ( \Phi A)) \times ( 1 \drop (\Phi A)) )\nonumber \\
+&= &<j\;k> \psi ((<i> \psi ( 2 \take (\Phi A))) \nonumber \\
+&&\;\;\;\;\;\times ( <i> \psi ( 1 \drop (\Phi A)))) \nonumber \\
+&= &<j\;k> \psi (( <i> \psi (\Phi A)) \times (<i+1> \psi (\Phi A))) \nonumber \\
+&=&<j\;k> \psi ((<((\rho A)[0] - 1 - i)> \psi A ) \nonumber \\
+&& \;\;\;\;\;\times  (<((\rho A)[0] - 1 -( i + 1))> \psi A ) )\nonumber \\
+&=&<j\;k> \psi ((< 2-i> \psi A ) \times (<1-i>  \psi A) )
+ \end{eqnarray}
+\normalsize
+This is the DNF, or semantic normal form. To proceed further we must know more about the data, i.e. layout, and architecture. 
+\subsection{Applying the Psi Correspondence Theorem\cite{jen94}]
+We now want to {\em transform} the DNF to a {\em Generic} Operational Normal Form (ONF) or way 
+to {\em build the code}.

### 2014-04-05 paper.tex
added s to make researcher plural
changes for psi reduction
Some more writing on the introduction of Pull

@@
 \cite{svenningsson2013combining}.
-
 In the array library Repa\cite{keller2010regular}, compiler supported
 rewrite rules are used to implement fusion \cite{jones2001playing}.
+
+Fusion is a powerful optimization which allows programming on a high
+level by composing functions without paying an performance penalty.
+
+Fusion effectively makes Pull arrays \emph{virtual} in that they will
+not be represented in memory. All intermediate arrays will be
+remove. But sometimes it is important to keep an array in memory, not
+least to produce the final result. For this reason, Pull arrays
+provide a function, variously known as \verb!force!, \verb!memorize!
+  or \verb!store!. We will refer to it as \verb!store! here. The
+  functional semantics of \verb!store! is the identity function,
+  i.e. it simply returns the same array as it was given. However, as a
+  side effect it stores the array to memory.
+
+A subtle point about \verb!store! is that the language is still pure
+with this function. The reason is that there is no way to observe from
+within the language whether an array is stored to memory or
+not. Therefore, the side-effect of \verb!store! cannot affect the
+value of a computation.
 
@@
 
+* Only one element type in MoA
+* Not much in the way of types in MoA
+
+* Shapes are arrays in MoA whereas they are a different type with Pull arrays
+
 \subsection{\(\psi\)-reductions and fusion}
 \label{sec:normalization}
+
+The similarity between MoA and Pull arrays is most clear when
+considering Operatonal Normal Form (ONF). Recall that ONF means that
 

### 2014-04-05 intromoa.tex
All notes in.
I hope you can read this today. I am  anxious to have your edits. Cut
or add anything you wish to make things clear.
I felt it would be easier to add all gory details and have you remove
as much as you can say in words during the talk. As mentioned I left
out many details.

@@
 This is the DNF, or semantic normal form. To proceed further we must know more about the data, i.e. layout, and architecture. 
-\subsection{Applying the Psi Correspondence Theorem\cite{jen94}]
-We now want to {\em transform} the DNF to a {\em Generic} Operational Normal Form (ONF) or way 
-to {\em build the code}.
+\subsection* {Applying the Psi Correspondence Theorem: PCT}
+We now want to {\em transform} the DNF above to a {\em Generic} Operational Normal Form (ONF) or way 
+to {\em build the code}.  The PCT~\cite{muljen94} turns cartesian coordinates into start, stops, and strides.
+Note that we still use {\em Psi-Reduction} but now just on vectors.  
+Let Y denote $(<2-i> \psi A ) \times ( <1-i> \psi A)$. Then the shape of Y, i.e.
+\begin{eqnarray}
+\rho Y= 1 \drop (\rho A) = 1 \drop <3\;5\;4> = <5\;4>
+\end{eqnarray}
+i.e. the bounds for j and k  above.
+Continuing with the PCT to create the {\em generic} ONF on p=2 processors, we use the index of the primary axis, i, $\ni\;\;0 \leq  i < p$, since $\rho X = <2\;4\;5>$.
+Since we are accessing contiguous planes we can collapse j and k into a single index l, $\ni \;\; 0 \leq l <20$.
+Using $\gamma_row$ for row major ordering, Y is transformed into:
+\small
+\begin{eqnarray}
+a[ (( \gamma(<2-i >; <1 \take (\rho A>))) \times \pi ( 1 \drop (\rho A))   )  + \iota \pi 1 \drop \rho A  ] \nonumber \\
+\times a[ (( \gamma(<1-i >; <1 \take (\rho A>))) \times \pi ( 1 \drop (\rho A))   )  + \iota \pi 1 \drop \rho A  ] 
+\end{eqnarray}
+Reducing, we get
+\begin{eqnarray}
+a[ (( \gamma(<2-i >; <3>))) \times \pi ( <5\;4>)  + \iota \pi <5\;4> ] \nonumber \\
+\times a[ (( \gamma(<1-i >; <3>))) \times \pi ( <5\;4>)   )  + \iota \pi <5\;4>  ]\nonumber \\
+a[( (2-i) \times  20)  + \iota 20 ] \nonumber \\
+\times a[( (i-1) \times 20 )  + \iota 20 ]
+\end{eqnarray}
+\normalsize
+Which breaks the indexing into start, stride, and stop, a universal hardware description and easiliy transformed
+to the mnemonics used at any hardware level ~\cite{fftharry}.
+We could add more dimensions to analyze prefetching, buffer sizes, vector registers, cache, etc.

### 2014-04-06 paper.tex
Flesh out the comparison between MoA and Pull arrays

@@
 University at Albany, SUNY}
-           {lenore@cs.albany.edu}
+           {lenore@albany.edu}
 
@@
 Having many different options to choose from can be a good thing but
-it can also lead to a balkanization of the community. When researcher
-dedicated to different formalisms can no longer talk to each other it
-inhibits collaborations and limits the impact of innovatation.
+it can also lead to a balkanization of the community. When researchers
+dedicated to different formalisms can no longer talk to each other,  it
+inhibits collaborations and limits the impact of innovation.
 
@@
 \label{sec:moa}
-
 \input{intromoa}
-
 \section{Overview of Pull arrays}
@@
 this function as an \emph{index function}. Additionally arrays also
-has a length.
+have a length.
 
@@
 
+In this section we relate MoA and Pull arrays.
+
+An initial observation is that MoA and Pull arrays where developed for
+different purposes and that can be seen already by looking at the
+notation. Mathematics of arrays have first and foremost been used for
+calculating array programs by hand. This is reflected in the short
+names and succinct notation which works very well when doing
+pencil-and-paper proofs.
+
+Pull arrays on the other hand where conveived in the context of a
+typed functional language. The notation is completely inherited from
+the syntax of Haskell \cite{marlow2010haskell}.
+
 \subsection{Types}
@@
 
-* Only one element type in MoA
-* Not much in the way of types in MoA
+When it comes to types MoA keeps things very simple. There are two
+types: arrays and scalars. Furthermore, scalars and singleton arrays
+are identified.
 
-* Shapes are arrays in MoA whereas they are a different type with Pull arrays
+Pull arrays inherits the type discipline from the language they are
+implemented in. When used in Haskell it means that they have a rich
+type system to utilize. First and foremost, Pull arrays are
+parametric, meaning that they can carry elements of any type. In
+particular the elements can be Pull arrays as another way of
+representing higher dimensional arrays. This representation is seldom
+used as it has poor performance.
 
-\subsection{\(\psi\)-reductions and fusion}
+Shapes in Pull arrays are represented by a completely different type,
+whereas in MoA shapes are simply arrays. Representing shapes as arrays
+is challenging in a typed language such as Haskell because it leads to
+a form of infinite recursion and while there are a number of way to
+break the recursion we have yet to find a way that does so and retains
+the efficiency and simplicity of Pull arrays.
+
+\subsection{DNF, \(\psi\)-reductions and fusion}
 \label{sec:normalization}
 
-The similarity between MoA and Pull arrays is most clear when
-considering Operatonal Normal Form (ONF). Recall that ONF means that
+The correspondence between MoA and Pull arrays is most clear when
+considering Denotaional Normal Form (DNF), \(\psi\)-reductions and
+fusion. Both formalisms define operations on arrays by how to index
+into the array. In MoA, the indexing function \(\psi\) is used,
+whereas for Pull arrays operations are defined as functions from index
+to the element at that index. These two forms are equivalent.
+
+Pull arrays are always kept on the form of an indexing function and a
+length. This representation is preserved for all operations. In MoA,
+arrays are treated as abstract during derivations. After calculations,
+\(\psi\)-reductions are performed to bring the expression into
+Denotational Normal Form. DNF is equivalent to Pull arrays, it shows
+how to compute an element from an array directly from an index.
+
+The process of applying \(\psi\)-reductions corresponds to performing
+fusion on Pull arrays. In both cases operations are inlined until
+there is a single expression for computing an element from an array
+given an index.
+
+Pull arrays use the function \verb!store! to prevent fusion and store
+an array in memory. MoA deals with memory differently. All arrays
+which are supposed to be allocated in memory are given a separate
+definition.
+
+The last step in calculating a program in MoA is to derive an
+Operational Normal Form which is explicit about how the array is laid
+out in memory. Pull arrays do not have anything corresponding to
+ONF. Instead, the decision of how the array is laid out is done by the
+compiler, which uses a pre-defined convention.
 
@@
 \hline
-δ & dim . extent
+δ & \verb!dim . extent!
 \\
-ρ & extent
+ρ & \verb!extent!
 \\
-ιξⁿ & -- generalized enumFromTo
+ιξⁿ & generalized \verb!enumFromTo!
 \\
-ψ & index
+ψ & generalized \verb!index!
 \\
@@
 \\
-γ & toIndex
+γ & \verb!toIndex!
 \\
-γ' & fromIndex
+γ' & \verb!fromIndex!
 \\
-$s \: \hat{ρ} \: ξ$ & reshape
+$s \: \hat{ρ} \: ξ$ & \verb!reshape!
 \\
-π x & product
+π x & \verb!product!
 \\
-τ & size
+τ & \verb!size!
 \\
@@
 \\
-ξ₁ f ξ₂ & zipWith f ξ₁ ξ₁
+ξ₁ f ξ₂ & \verb!zipWith f! ξ₁ ξ₁
 \\
-σ f ξ & map f ξ
+σ f ξ & \verb!map f! ξ
 \\
-ξ f σ & map f ξ
+ξ f σ & \verb!map f! ξ
 \\
-$\bigtriangleup$ & take
+$\bigtriangleup$ & \verb!take!
 \\
-$\bigtriangledown$ & drop
+$\bigtriangledown$ & \verb!drop!
 \\
-$op^{red}$ & fold
+$op^{red}$ & \verb!fold op!
 \\
-Φ & reverse
+Φ & \verb!reverse!
 \\
-Θ & rotate
+Θ & \verb!rotate!
 \\
-$\varobslash$ & -- generalized transpose
+$\varobslash$ & generalized \verb!transpose!
 \\

### 2014-04-06 paper.tex
editing

@@
 new expressive notion of higher dimensionality for Push arrays
-\cite{keller2010regular}. The number dimensions of an array, referred
+\cite{keller2010regular}. The number of dimensions of an array, referred
 to as the \emph{shape}, are determined statically in the type
@@
 be overly constraining and exclude useful programs. A key innovation in
-Repa is the notion on \emph{shape polymorphism} which makes it
+Repa is the notion of \emph{shape polymorphism} which makes it
 possible to write functions which accepts arrays of any shape. Below

### 2014-04-06 paper.tex
please read email
please read my notes. I'll await your edits.

@@
 As an example of fusion consider the example
+\newline
 \verb!map (*2) (map (*5) arr)!, where each element of the array
@@
 The correctness of fusion relies crucially on the language being
-\emph{pure}. If side effects where allowed in the indexing function
+\emph{pure}. If side effects were allowed in the indexing function
 fusion would not be correct.
@@
 not be represented in memory. All intermediate arrays will be
-remove. But sometimes it is important to keep an array in memory, not
+removed. But sometimes it is important to keep an array in memory, not
 least to produce the final result. For this reason, Pull arrays
@@
 
-An initial observation is that MoA and Pull arrays where developed for
+An initial observation is that MoA and Pull arrays were developed for
 different purposes and that can be seen already by looking at the
-notation. Mathematics of arrays have first and foremost been used for
+notation. Mathematics of Arrays has first and foremost been used for
 calculating array programs by hand. This is reflected in the short
@@
 
-Pull arrays on the other hand where conveived in the context of a
+Pull arrays on the other hand were conveived in the context of a
 typed functional language. The notation is completely inherited from
@@
 are identified.
+% scalars are arrays from an MoA and Psi Calculus perspective.
 

### 2014-04-07 paper.tex
Type updated
Talking may be helpful if my emails seem confusing.

@@
 
-When it comes to types MoA keeps things very simple. There are two
-types: arrays and scalars. Furthermore, scalars and singleton arrays
-are identified.
+When it comes to types MoA keeps things very simple. There is just one
+type, arrays.  Scalars are arrays;  their shape is the empty vector. Consequently, we have the identify given any array, $\xi$
+\[( \iota (\rho \xi)  )\psi \xi \equiv \xi\]
+That is, given an array's shape, all indices can be generated to index the entire array.
 % scalars are arrays from an MoA and Psi Calculus perspective.

### 2014-04-08 intromoa.tex
small typo change
updates
Check it out and let me know if I transition well to ONF.

@@
 
-MoA, an acoronym for {\em A Mathematics of Arrays\cite{thesis}}, is a theory of arrays with both an algebra and a calculus of 
+MoA, an acoronym for {\em A Mathematics of Arrays\cite{thesis}}, is a theory of arrays with both an 
+algebra and a calculus of 
 indexing based on shapes, i.e. sizes of each dimension. The index calculus, refered to as {\em The Psi Calculus}, 
@@
 calculus\cite{tu2,tu3}. This awareness led to Mullin's collaborations with Klaus Berkling to combine MoA, Psi Calculus, and
-the Lambda Lalculus, theoretically an ideal enviornment\cite{berkling} and basis for a 
+the Lambda Calculus, theoretically an ideal enviornment\cite{berkling} and basis for a 
 functional programming language with 
@@
 && \;\;\;\;\;\times  (<((\rho A)[0] - 1 -( i + 1))> \psi A ) )\nonumber \\
-&=&<j\;k> \psi ((< 2-i> \psi A ) \times (<1-i>  \psi A) )
+&=&<j\;k> \psi ((< 2-i> \psi A ) \times (<1-i>  \psi A) ) \nonumber \\
+&=& (< (2-i)\; j\;k> \psi A )  \times (<(i-i) \; j\;k> \psi A ) 
  \end{eqnarray}
@@
 We now want to {\em transform} the DNF above to a {\em Generic} Operational Normal Form (ONF) or way 
-to {\em build the code}.  The PCT~\cite{muljen94} turns cartesian coordinates into start, stops, and strides.
+to {\em build the code}.  The PCT~\cite{muljen94}, which generalizes the mapping function 
+gamma, $\gamma$,   turns short indices into start, stops, and strides.
 Note that we still use {\em Psi-Reduction} but now just on vectors.  
-Let Y denote $(<2-i> \psi A ) \times ( <1-i> \psi A)$. Then the shape of Y, i.e.
+We start by analyzing the shape of the result in conjunction with the indicies involved
+in the computation and observe that the indices for $j$ and $k$ are contiguous allowing
+us to only use the primary axis index in the mapping.
+Thus, let Y denote $(<2-i> \psi A ) \times ( <1-i> \psi A)$. Then the shape of Y, i.e.
 \begin{eqnarray}
@@
 \end{eqnarray}
-i.e. the bounds for j and k  above.
+i.e. the bounds for $j$ and $k$  above.
 Continuing with the PCT to create the {\em generic} ONF on p=2 processors, we use the index of the primary axis, i, $\ni\;\;0 \leq  i < p$, since $\rho X = <2\;4\;5>$.
-Since we are accessing contiguous planes we can collapse j and k into a single index l, $\ni \;\; 0 \leq l <20$.
-Using $\gamma_row$ for row major ordering, Y is transformed into:
+Using $\gamma_{row}$ for row major ordering, Y is transformed into:
 \small
@@
 \normalsize
-Which breaks the indexing into start, stride, and stop, a universal hardware description and easiliy transformed
+which transforms the indexing into start, stride, and stop, a universal hardware description and easiliy transformed
 to the mnemonics used at any hardware level ~\cite{fftharry}.

### 2014-04-08 intromoa.tex
a few more updates for clarity
I added a few more sentences in the ONF paragraph to assist in clarity.
Again, take out whatever you wish.

@@
 to the mnemonics used at any hardware level ~\cite{fftharry}.
-We could add more dimensions to analyze prefetching, buffer sizes, vector registers, cache, etc.
+
+The  {\em iota} operation depicts  20 sequential 
+accesses. If we wanted to map these operations to other than  scalar registers
+we could conceptually add more dimensions to analyze prefetching, buffer sizes, vector registers, caches, etc.
+Suppose in this example we have two vector registers to do the addition, both with length 4, we could
+abstract this operation into a 5 by 4 where we can now analyze the cost of 5 vector register loads and
+stores and determine if that is more cost effective than 20 scalar operations. Thus prior to code generation we
+{\em could} analyze which menomnic to map to from the ONF.

### 2014-04-08 intromoa.tex
more edits
More edits to make things flow better.

@@
 We now want to {\em transform} the DNF above to a {\em Generic} Operational Normal Form (ONF) or way 
-to {\em build the code}.  The PCT~\cite{muljen94}, which generalizes the mapping function 
-gamma, $\gamma$,   turns short indices into start, stops, and strides.
+to {\em build the code}.  Recall that $\gamma$ maps a {\em full} index , $\vec i$, of an arbitrary array $A$ to an 
+offset from the start of A, $@A$ in memory,  denoted by $a[\gamma(\vec i;(\rho A))]$.
+The PCT~\cite{muljen94},  generalizes the mapping function 
+gamma, $\gamma$,   and turns {\em short} indices into start, stops, and strides.
 Note that we still use {\em Psi-Reduction} but now just on vectors.  
@@
 in the computation and observe that the indices for $j$ and $k$ are contiguous allowing
-us to only use the primary axis index in the mapping.
+us to only use the primary axis index, a short index, in the mapping.
 Thus, let Y denote $(<2-i> \psi A ) \times ( <1-i> \psi A)$. Then the shape of Y, i.e.

### 2014-04-09 intromoa.tex
more edits
more edits for clarity

@@
         &   = & <1 \; 2> \psi A  \nonumber \\
+&=&\mbox {Thus,} \;\;\forall \;\; j \;\;\ni \;\;0 \leq j < 4 \nonumber \\
+&=& = <j>  \psi <1 \; 2> \psi A  \nonumber \\
+&=& <1\:2\;j\;> \psi A \nonumber\\
 \end{eqnarray}
-This process of simplifying the expression for the item in terms of
+
+This process of simplifying expression (1) in terms of
 its Cartesian coordinates is called {\em Psi Reduction}. The operations of MoA
 have been designed so that all expressions can be reduced to a minimal
-normal form~\cite{mul88}.
+normal form~\cite{mul88}, the DNF.
 Some MoA operations defined by {\em psi} are found in Fig.\ref{moa}.

### 2014-04-10 intromoa.tex
removed a table
fixed more typos and removed Figure 1

--- 2014-04-08 intromoa.tex
@@
         &   = & <1 \; 2> \psi A  \nonumber \\
+\mbox {Thus,} \;\;\forall \;\; j \;\;\ni \;\;0 \leq j < 4 \nonumber \\
+&=&  <j>  \psi <1 \; 2> \psi A  \nonumber \\
+&=& <1\:2\;j\;> \psi A \nonumber\\
 \end{eqnarray}
-This process of simplifying the expression for the item in terms of
+
+This process of simplifying expression (1) in terms of
 its Cartesian coordinates is called {\em Psi Reduction}. The operations of MoA
 have been designed so that all expressions can be reduced to a minimal
-normal form~\cite{mul88}.
-Some MoA operations defined by {\em psi} are found in Fig.\ref{moa}.
-
-%\break
-\begin{figure}[h]
-%\scriptsize
-\tiny
-\begin{tabular}{|l|l|l|}
-\hline\hline
-{\bf Symbol} & {\bf Name}& {\bf Description} \\ \hline
-$\delta$ & Dimensionality & Returns the number of dimensions of an array.\\ \hline
-$\rho$ & Shape & Returns a vector of the upper bounds \\
-& &or sizes of each dimension in an array. \\ \hline
-$\iota \xi^n$  & Iota & When $n=0$(scalar),
-returns a vector containing elements\\
-& & $0$, to $\xi^0 - 1$. When $n = 1$(vector), returns an \\
-& & array of indices  defined by the shape vector $\xi^1$ \\ \hline
-$\psi$ & Psi & The main indexing function of the Psi Calculus \\
-& & which defines all operations in MoA.   Returns a scalar \\
-& & if a full index is provided, a sub-array otherwise. \\ \hline
-\mbox{rav} & Ravel & vectorizes a multi-dimensional array based   \\
-&&on an  array's layout($\gamma_{row},\gamma_{col},\gamma_{sparse}, ...$)\\ \hline
-$\gamma$ & Gamma & Translates indices into offsets given a shape. \\ \hline
-$\gamma^{'}$ & Gamma Inverse & Translates offsets into indices given a shape. \\ \hline
-$\vec s \rshp \xi $ & Reshape& Changes the shape vector of an array,
- possibly affecting  \\
-& &its dimensionality. Reshape depends on
-layout($\gamma$). \\ \hline
-$\pi \vec x$ & Pi & Returns a scalar and is equivalent to $\prod_{i=0}^{(\tau x)-1}x[i]$ \\ \hline
-$\tau$& Tau & Returns the number of components in an array,($\tau \xi \equiv
-\pi (\rho \xi) $) \\ \hline
-$\xi_l \cat \xi_r $ & Catenate& Concatenates two arrays over their
-primary axis. \\ \hline
-$\xi_l f \xi_r $ & Point-wise  & A data parallel application of $f$ is performed \\
-&Extension &  between all elements of the arrays.\\ \hline
-$\sigma f \xi_r $ & Scalar Extension & $\sigma$ is used with
-every component of $\xi_r$ in the data parallel \\
-$\xi_l f \sigma$ & &application of $f$. \\ \hline
-$\take$&Take& Returns a sub-array from the beginning or end of an array\\
-& & based on its argument being positive or negative. \\ \hline
-$\drop$ & Drop & The inverse of Take \\ \hline
-$_{op} \mbox{red}$&Reduce& Reduce an array's dimension by one by applying \\
-& & op over the primary axis of an array. \\ \hline
-$\Phi$& Reverse& Reverses the components of an array. \\ \hline
-$\Theta $& Rotate & Rotates, or shifts cyclically, components of an
-array. \\ \hline
-$\transpose$ & Transpose & Transposes the elements of an array based on \\
-&&a given permutation vector \\ \hline
-$\Omega$& Omega& Applies a unary or binary function to array argument(s) \\
-&&given partitioning information. $\Omega$ is used to perform all operations \\&& (defined over the primary axis only) over all dimensions. \\ \hline
-\end{tabular}
-\caption{\label{moa}Summary of MoA Operations}
-\end{figure}
+normal form~\cite{mul88}, the DNF.
 

### 2014-04-11 intromoa.tex
Various changes, writing a bit on Conclusions and some shortening

@@
 indexing based on shapes, i.e. sizes of each dimension. The index calculus, refered to as {\em The Psi Calculus}, 
-gets its names from the {\em Psi}  function which is the foundational function in the theory. The name {\em Psi} 
-was inspired by Dirac's {\em Psi Vectors} and his communications with Einstein to have a mathematical way to reason
-about them. These results began with the awareness that Iverson's APL\cite{iverson62}
+gets its names from the {\em Psi}  function which is the foundational function in the theory. These results began with the awareness that Iverson's APL\cite{iverson62}
 was inspired by Sylvester's Universal Algebra\cite{sylvester} and the fact that he and Cayley invented matrices.
-The programming language, APL, embodied Ken's notation into an interpretive computing environment. It was at this point
-that theoretical anamolies were introduced.
-
-Not only did  Ken believe that APL's array algebra was the ideal basis for a programming language, he (and others) believed that
-an ideal machine language supported array operations. This led to HP building an 
-APL machine based on Phil Abram's dissertation\cite{abrams70}. Abrams was the first to identify ways 
-to define an algebra of arrays based on shapes; others attempted to
-put closure on his work\cite{lyon,guibas}. Abrams believed that if array operations were based on 
-indexing with  shapes, all  indexing in the array expression could be  composed
-and  thus could minimize the creation (and deletion) of temporary arrays,
-a huge overhead. One of the reasons HP's APL Machine did not succeed was that this analysis was done at run time.
-It was  Alan Perlis, and later others\cite{miller,budd84},  who realized that these optimizations should be done at compile time.
-Attempts at compilation led Perlis to realize that  anomalies in the language~\cite{gerhart}
-made it difficult to optimize and impossible to verify correctness of 
-programs leading him to recommend a functional array calculator based on the lambda 
-calculus\cite{tu2,tu3}. This awareness led to Mullin's collaborations with Klaus Berkling to combine MoA, Psi Calculus, and
-the Lambda Calculus, theoretically an ideal enviornment\cite{berkling} and basis for a 
-functional programming language with 
-arrays\cite{mul91,budd91,hm92,mul93,mul94,mul96}.
 

### 2014-04-11 paper.tex
Set the title
Pluralization
Various changes, writing a bit on Conclusions and some shortening

@@
 
-\title{Title Text}
-\subtitle{Subtitle Text, if any}
+\title{Relating two Abstractions for Formal Array Languages}
+%\subtitle{Subtitle Text, if any}
 
@@
   \ref{sec:types}), fusion/\(\psi\)-reduction (section
-  \ref{sec:normalization}), how they deal with higher dimensions
-  (section \ref{sec:highdim}) and notation (section \ref{sec:notation}).
+  \ref{sec:normalization}), 
+  how they deal with higher dimensions (section \ref{sec:highdim})
+  and notation (section \ref{sec:notation}).
 \end{itemize}
@@
 \cite{keller2010regular,Axelsson:2010:Feldspar,Mainland:2010:Nikola,Svensson:2011:Obsidian,Claessen:2012:Expressive,Ankner:2013:AnEDSL,lippmeier2011efficient}. One
-of the central abstractions in this line of work is the \emph{pull
-  array}, also know as delayed array. The can be defined as follows
+of the central abstractions in this line of work is \emph{pull
+ arrays}, also know as delayed arrays. The can be defined as follows
 (we use Haskell \cite{marlow2010haskell} to demonstrate functions
@@
 \end{itemize}
+
+Figure \ref{fig:pullex} shows some examples of functions on Pull
+arrays. Things to note are that most functions are polymorphic; they
+can operate on arrays of any type of element.
 
@@
   = Pull (\i -> ixf ((i + k) `mod` l)) l
+
+enumFromTo :: Int -> Int -> Pull Int
+enumFromTo f t = Pull (\i -> i + f) (t - f + 1)
 \end{verbatim}
 \caption{Example functions on pull arrays}
+\label{fig:pullex}
 \end{figure}
@@
 
+Below we show how higher dimensional Pull arrays can be defined using
+the \verb!Shape! type. We also show some functions adapted for the new
+representation.
 \begin{verbatim}
@@
 When it comes to types MoA keeps things very simple. There is just one
-type, arrays.  Scalars are arrays;  their shape is the empty vector. Consequently, we have the identify given any array, $\xi$
+type, arrays.  Scalars are arrays;  their shape is the empty vector. Consequently, we have the identity given any array, $\xi$
 \[( \iota (\rho \xi)  )\psi \xi \equiv \xi\]
@@
 used as it has poor performance.
-
-Shapes in Pull arrays are represented by a completely different type,
-whereas in MoA shapes are simply arrays. Representing shapes as arrays
-is challenging in a typed language such as Haskell because it leads to
-a form of infinite recursion and while there are a number of way to
-break the recursion we have yet to find a way that does so and retains
-the efficiency and simplicity of Pull arrays.
 
@@
 \label{sec:highdim}
+
+Shapes in Pull arrays are represented by a completely different type,
+whereas in MoA shapes are simply arrays. Representing shapes as arrays
+is challenging in a typed language such as Haskell because it leads to
+a form of infinite recursion and while there are a number of way to
+break the recursion we have yet to find a way that does so and retains
+the efficiency and simplicity of Pull arrays.
+
+Apart from the differences in the type system, MoA and Pull arrays
+treat shapes relatively similar. The type of shapes in Pull arrays can
+be seen as just another representation of arrays, one where the length
+is part of the type.
 
@@
 
-\section{Related work}
+\section{Conclusion and Future work}
 
-\section{Future work}
+Mathematics of Arrays and Pull arrays are a lot more closely related
+than what they first might appear. In this paper we have outlined the
+relationship between fusion and \(\psi\)-reduction, Pull arrays and
+Denotational Normal Form, and the various types of operations used.
+Our hope is that this paper will help researchers from both
+communities interact and learn from each other.
 
-\section{Conclusion}
+We have identified a few points where MoA and Pull arrays can learn
+from each other. The treatment of higher dimensions in MoA is more
+thorough and there are operations like $\Omega$ which haven't appeared
+in the literature on Pull arrays.
 
-\acks
+Pull arrays has a dual representation, Push arrays
+\cite{Claessen:2012:Expressive}. It can deal with some of the
+shortcomings of Pull arrays, in particular more efficient
+concatenation and sharing of computations between array elements. It
+would be interesting to see how the ideas of Push array carries over
+to MoA.
 
-Acknowledgments.
+%% \acks
+
+%% Acknowledgments.
 

### 2014-04-11 intromoa.tex
Shorten the paper so that we're within 6 pages

@@
 \[                  <i \; j \; k> \]
-denotes a vector of length three. $<>$ will denote the empty vector which is 
-also sometimes written as $\Theta$.
+denotes a vector of length three. $<>$ will denote the empty vector.
 
@@
 \> \> \>and when $k<0$ the vector of the first $t_u - |k|$ items of $u$ \\
-\>$k \;\theta \;u$ \>  \>  when $k \geq 0$  the vector of $(k  \drop  u) \cat ( k \take u ) $\\
-\> \> \>and when $k<0$ the vector or $(k \take u ) \cat ( k \drop u ) $
 \end{tabbing}
@@
 also that all operations are over the 0th axis. The operator $\Omega$~\cite{mul88} extends operations over all other dimensions.
-\subsubsection*{Example}
-Consider the evaluation of the following expression using the
-3 by 5 by 4 array, $A$, introduced in Section~\ref{indshp}.
-\begin{eqnarray}
-             <1\;\; 2> \psi (2 \take \Phi A) 
-\end{eqnarray}
-where A is the array given in the previous section.
-The shape of the result is 
-\begin{eqnarray}
-                & &  2 \drop s_{(2 \take \Phi A)} \nonumber \\
-                & = & 2 \drop (<2> \cat (1 \drop s_{\Phi A} )) \nonumber \\
-               & = & 2 \drop (<2> \cat (1 \drop s_A)) \nonumber \\
-               & = & 2 \drop (<2> \cat <5\; 4>) \nonumber \\
-               & = & 2 \drop <2\; 5 \;4> \nonumber \\
-              &  = &<4> . 
-\end{eqnarray}
-The expression can be simplified using the definitions:
-\begin{eqnarray}
-          &&   <1\; 2> \psi (2 \take \Phi A) \nonumber \\
-           & = &  <1 \; 2> \psi \Phi A \nonumber \\
-         &   = & <2> \psi (<1> \psi \Phi A) \nonumber \\
-         &  = & <2> \psi (<3 - 1 - 1 > \psi A) \nonumber \\
-        &   = & <1 \; 2> \psi A  \nonumber \\
-\mbox {Thus,} \;\;\forall \;\; j \;\;\ni \;\;0 \leq j < 4 \nonumber \\
-&=&  <j>  \psi <1 \; 2> \psi A  \nonumber \\
-&=& <1\:2\;j\;> \psi A \nonumber\\
-\end{eqnarray}
-
-This process of simplifying expression (1) in terms of
-its Cartesian coordinates is called {\em Psi Reduction}. The operations of MoA
-have been designed so that all expressions can be reduced to a minimal
-normal form~\cite{mul88}, the DNF.
 

### 2014-04-11 paper.tex
Shorten the paper so that we're within 6 pages

@@
 We have identified a few points where MoA and Pull arrays can learn
-from each other. The treatment of higher dimensions in MoA is more
-thorough and there are operations like $\Omega$ which haven't appeared
-in the literature on Pull arrays.
+from each other. The treatment of higher dimensions in MoA is in
+general more thorough than Pull arrays. There are operations like
+$\Omega$ which haven't appeared in the literature on Pull arrays but
+which would provide a very useful addition.
 
-Pull arrays has a dual representation, Push arrays
+Pull arrays has a dual representation: Push arrays
 \cite{Claessen:2012:Expressive}. It can deal with some of the

### 2014-04-11 paper.tex
looks good
I made one typo change and added a footnote about materialization. Let
me know what else we should do.

@@
 Fusion is a powerful optimization which allows programming on a high
-level by composing functions without paying an performance penalty.
+level by composing functions without paying a performance penalty.
 
@@
 not be represented in memory. All intermediate arrays will be
-removed. But sometimes it is important to keep an array in memory, not
-least to produce the final result. For this reason, Pull arrays
+removed. But sometimes it is important to keep an array in memory
+\footnote{Example: Indexing a huge transposed array that has a few rows and many columns. Non-materialization
+may cause many cache misses and page faults.}, 
+not least to produce the final result. For this reason, Pull arrays
 provide a function, variously known as \verb!force!, \verb!memorize!

### 2014-04-11 paper.tex
Create a figure for the rosetta table
Wibble
Removed a statement about MoA types as it used undefined functions

@@
 When it comes to types MoA keeps things very simple. There is just one
-type, arrays.  Scalars are arrays;  their shape is the empty vector. Consequently, we have the identity given any array, $\xi$
-\[( \iota (\rho \xi)  )\psi \xi \equiv \xi\]
-That is, given an array's shape, all indices can be generated to index the entire array.
+type, arrays.  Scalars are arrays;  their shape is the empty vector.
+
+% I've removed this statement because \rho is not defined in the paper
+%% Consequently, we have the identity given any array, $\xi$
+%% \[( \iota (\rho \xi)  )\psi \xi \equiv \xi\]
+%% That is, given an array's shape, all indices can be generated to index the entire array.
 % scalars are arrays from an MoA and Psi Calculus perspective.
@@
 
+Figure \ref{fig:rosetta} gives a translation between function names
+used in MoA and Pull arrays.
+
+\begin{figure}
+\begin{center}
 \begin{tabular}{|@{}l|l@{}|}
@@
 \end{tabular}
+\end{center}
+\caption{A correspondence between functions in MoA and Pull arrays}
+\label{fig:rosetta}
+\end{figure}
 

### 2014-04-11 paper.tex
Explaining the difference between various functions in MoA and Pull arrays

@@
 Figure \ref{fig:rosetta} gives a translation between function names
-used in MoA and Pull arrays.
+used in MoA and Pull arrays. The intention of this figure is to work
+as a rosetta stone, helping to translate from one formalism to the
+other.
+
+As can be seen in the figure, not all operations in MoA have a
+corresponding function in the Pull array literature. Some of these
+functions are missing because Pull arrays do not have a notion
+corresponding to ONF in MoA. In particular, the function $rav$ is used
+in ONF to decide on the particular layout of the array, and this step
+is implicit for Pull arrays. However, $\Omega$ is a good candidate to
+transfer to the formalism of Pull arrays.
+
+The difference between ψ in MoA and \verb!index! i Pull arrays is that
+ψ can be used with partial indexes, a central notion in MoA. Pull
+arrays solve this problem in a more roundabout way, using a notion of
+\emph{slices} \cite{keller2010regular}. We leave out the explanation
+of slices for reasons of space.
 
@@
 \\
-ιξⁿ & generalized \verb!enumFromTo!
+ιξⁿ & \verb!enumFromTo!
 \\
@@
 \\
-rav & -- flattens to a one-dimensional array
+rav & -- No corresponding function
 \\
@@
 \\
-$f Ω_d ξ$ & -- Apply $f$ along the dimension $d$ of $ξ$
+$f Ω_d ξ$ & -- No corresponding function
 \\
@@
 \end{figure}
-
-\subsection{Example programs}
 

### 2014-04-11 paper.tex
Removed some comments in the code to save space. More text about shapes

@@
 \begin{verbatim}
--- Mapping a function over an array
 map :: (a -> b) -> Pull a -> Pull b
@@
 
--- Apply a binary operation itemwise on two arrays
 zipWith :: (a -> b -> c)
@@
 
--- Takes the n first elements of an array
 take :: Int -> Pull a -> Pull a
@@
 
--- Drops the n first elements of an array
 drop :: Int -> Pull a -> Pull a
@@
 
--- Rotate an array k steps to the left
 rotate :: Int -> Pull a -> Pull a
@@
   = Pull (\i -> ixf ((i + k) `mod` l)) l
+
+reverse :: Pull a -> Pull a
+reverse (Pull ixf l) = Pull (\i -> ixf (l - i - 1) l
 
@@
 
-There are two different ways of implementing Pull arrays in the
-literature. The simplest one comes from using Pull arrays in an
-embedded langauge, i.e. when the array language is implemented as a
-library in another host language. Examples of such languages are
-Pan\cite{elliott2003compiling}, Feldspar\cite{Axelsson:2010:Feldspar},
-Obsidian\cite{Svensson:2011:Obsidian} and
-Nikola\cite{Mainland:2010:Nikola}. By using the technique of
-implementing Pull arrays as \emph{shallow embeddings} on top of a
-\emph{deeply embedded} code language, fusion comes completely for free
-\cite{svenningsson2013combining}.
-In the array library Repa\cite{keller2010regular}, compiler supported
-rewrite rules are used to implement fusion \cite{jones2001playing}.
-
-Fusion is a powerful optimization which allows programming on a high
-level by composing functions without paying a performance penalty.
-
 Fusion effectively makes Pull arrays \emph{virtual} in that they will
@@
 value of a computation.
+
+\subsection{Implementing Fusion}
+
+There are two different ways of implementing Pull arrays in the
+literature. The simplest one comes from using Pull arrays in an
+embedded langauge, i.e. when the array language is implemented as a
+library in another host language. Examples of such languages are
+Pan\cite{elliott2003compiling}, Feldspar\cite{Axelsson:2010:Feldspar},
+Obsidian\cite{Svensson:2011:Obsidian} and
+Nikola\cite{Mainland:2010:Nikola}. By using the technique of
+implementing Pull arrays as \emph{shallow embeddings} on top of a
+\emph{deeply embedded} code language, fusion comes completely for free
+\cite{svenningsson2013combining}.
+In the array library Repa\cite{keller2010regular}, compiler supported
+rewrite rules are used to implement fusion \cite{jones2001playing}.
 
@@
 the \verb!Shape! type. We also show some functions adapted for the new
-representation.
+representation. Note the shape polymorphism in the definitions of
+\verb!map! and \verb!zipWith!. They can work with arrays of any
+shapes, as long as they are the same shape.  The function
+\verb!reverse! is particularly noteworthy because the type reflects
+that the input and output arrays need to have at least one dimension.
+
 \begin{verbatim}
@@
   = Pull (\sh -> f (ixf1 sh) (ixf2 sh2))
+
+reverse :: Pull (sh :. Data Length) a ->
+           Pull (sh :. Data Length) a
+reverse (Pull ixf sh@(_ :. l)) =
+ Pull (\(sh' :. i) -> ixf (sh' :. (l - i - 1))) sh
 \end{verbatim}

### 2014-04-11 intromoa.tex
squeezing out text
I have edited as much as I can. Defs are essential.
If you can find anything else please go ahead and edit.
I know it is getting late for you and we need to get this in.
Let me know when you submit.
Lenore

@@
 normal form (Denotational Normal Form-DNF) then to an Operational Normal Form 
-(ONF). The ONF describes how to {\em build} the code using 
-{\em start, stop, stride}, and {\em count}: a universal machine abstraction.
-This is possible by {\em abstractly} lifting the dimension of an expression to include all aspects
-of a processor, memory, communication topology. For example, to map a 2-d array expression to
-n  processors we would lift the dimenion to 3-d. 
+(ONF). 
 %
@@
 
-Most of the common array manipulation operations found in languages like Fortran90, Matlab, ZPL, etc.,
- can be defined from $\psi$ and a few elementary vector operations.
-
-	We now introduce notation to permit us to define $\psi$ formally and to develop the
-{\em Psi Correspondence Theorem}~\cite{muljen94}, which is central to the effective exploitation of MoA in array
-computations. We will use $A, B, ...$ to denote an array of numbers (integer, real, or complex).
+	We now introduce notation to permit us to define $\psi$ formally. We will use $A, B, ...$ to denote an array of numbers (integer, real, or complex).
 An array's dimensionality will be denoted by $d_A$ and will be assumed to 
@@
 	The precise mapping of $A$ to $a$ is determined by a one-to-one 
-ordering function, $gamma$.  Although the choice of ordering is arbitrary, it is 
-essential in the following that a specific one be assumed. By convention 
-we assume the items of $A$ are placed in $a$ according to the lexicographic 
-ordering of the indices of $A$. This is often referred to 
-as $row \; major \; ordering$. 
-Many programming languages lay out the items of multidimensional
-arrays in memory in a contiguous segment using this ordering. 
-Fortran uses the ordering corresponding to a transposed array in which the 
-axes are reversed {\em column major}. Scalars are introduced as arrays with an empty shape vector.
+ordering function, $gamma$
+\footnote{There are a family of gamma functions: 
+$\gamma_{row}, \gamma_{column}, \gamma_{regular sparse}$, etc. In this case we assume row major ordering.}.  
+Scalars are introduced as arrays with an empty shape vector.
 
@@
 \>$\pi \; v$ \> \>a scalar which is the product of the components of v \\
-\>$k \;\take \;u$ \>\>	when $k \geq 0$ the vector of the first $k$ items of $u$,(called {\em take}) \\
+\>$k \;\take \;u$ \>\>	when $k \geq 0$ the vector of the first $k$ items of $u$) \\
 \> \> \>and when $k<0$ the vector of the last $k$ items of $u$ \\
-\>$k \;\drop \;u$ \>  \> when $k \geq 0$ the vector of $t_u - k$ last items of $u$,(called{ \em drop}) \\
+\>$k \;\drop \;u$ \>  \> when $k \geq 0$ the vector of $t_u - k$ last items of $u$) \\
 \> \> \>and when $k<0$ the vector of the first $t_u - |k|$ items of $u$ \\

### 2014-04-11 paper.tex
Use the preprint option to the document class Also remove keywords and classifications
Remove the mention of deforestation
Change fusion example to use same as for MoA

@@
 
-\documentclass{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
@@
 \usepackage[utf8x]{inputenc}
+\usepackage{fancyvrb}
 
@@
 
-\category{CR-number}{subcategory}{third-level}
+% \category{CR-number}{subcategory}{third-level}
 
@@
 % you may leave them out
-\terms
-term1, term2
+%\terms
+%term1, term2
 
-\keywords
-keyword1, keyword2
+%\keywords
+%Arrays
 
@@
 
-\begin{verbatim}
+\begin{small}
+\begin{Verbatim}
 data Pull a = Pull { ixf    :: (Int -> a)
@@
                    }
-\end{verbatim}
+\end{Verbatim}
+\end{small}
 
@@
 \begin{figure}
-\begin{verbatim}
+\begin{small}
+\begin{Verbatim}
 map :: (a -> b) -> Pull a -> Pull b
@@
 enumFromTo f t = Pull (\i -> i + f) (t - f + 1)
-\end{verbatim}
+\end{Verbatim}
+\end{small}
 \caption{Example functions on pull arrays}
@@
 fusion\cite{gill1993short,axelsson2010feldspar,keller2010regular}. Fusion
-is a program transformation which removes intermediate data structures
-and is a decendant of deforestation
-\cite{wadler1990deforestation}. Fusion supercedes transformations such
+is a program transformation which removes intermediate data structures.
+Fusion supercedes transformations such
 as loop fusion in optimizing compilers for imperative languages.  Pull
@@
 
-As an example of fusion consider the example
-\newline
-\verb!map (*2) (map (*5) arr)!, where each element of the array
-\verb!arr! is first multiplied by five and then again by
-two. Conceptually, there is an intermediate array created after the
-first \verb!map! but thanks to fusion it will be removed. Fusion will proceed
-as follows:
+As an example of fusion consider the example from the section which
+was used to illustrate $\psi$-reductions. Here is how it is written
+using Pull arrays for the one-dimensional case:\\
+{\small
+\verb!zipWith (*) (take 2 (reverse a)) (drop 1 (reverse a))!}.\\
+Conceptually there are many intermediate arrays here, but they will be
+removed by fusion. Fusion will proceed by inlining the definition
+of the functions as follows:
 
-\begin{verbatim}
-map (*2) (map (*5) arr)
+\begin{small}
+\begin{Verbatim}
+zipWith (*) (take 2 (reverse a))
+            (drop 1 (reverse a))
   => { arr }
-map (*2) (map (*5) (Pull ixf l)
-  => { map }
-map (*2) (Pull ((*5) . ixf) l)
-  => { map }
-Pull ((*2) . (*5) . ixf) l)
-\end{verbatim}
+zipWith (*) (take 2 (reverse (Pull ixf l))
+            (drop 2 (reverse (Pull ixf l)))
+  => { reverse * 2 }
+zipWith (*) (take 2 (Pull (\i -> ixf (l - i - 1)) l))
+            (drop 1 (Pull (\i -> ixf (l - i - 1)) l))
+  => { take , drop }
+zipWith (*) (Pull (\i -> ixf (l - i - 1)) (min 2 l))
+            (Pull (\i -> ixf (l - i)) (max (l - 1) 0))
+  => { zipWith }
+Pull (\i -> ixf (l - i - 1) * ixf (l - i))
+  (min (min 2 l) (max (l - 1) 0))
+\end{Verbatim}
+\end{small}
 
-The result is a single Pull array, the intermediate array has been removed.
+The result is a single Pull array, the intermediate array has been
+removed. The result is the same in MoA, modulo syntax and the fact
+that we here use a one-dimensional array of unknown size.
 

### 2014-04-11 paper.tex
Squeeze some more by using smaller verbatim font
Spelling

@@
 history of computing. For this reason we have seen a plethora of
-formalisms and languages being developed to aid theoriticians and
+formalisms and languages being developed to aid theoreticians and
 developers to produce more suitable abstractions and faster programs.
-Examples include Fortran, APL and Matlab but the list goes on and on.
+Examples include FORTRAN, APL and Matlab but the list goes on and on.
 Having many different options to choose from can be a good thing but
@@
 is a program transformation which removes intermediate data structures.
-Fusion supercedes transformations such
+Fusion supersedes transformations such
 as loop fusion in optimizing compilers for imperative languages.  Pull
@@
 literature. The simplest one comes from using Pull arrays in an
-embedded langauge, i.e. when the array language is implemented as a
+embedded language, i.e. when the array language is implemented as a
 library in another host language. Examples of such languages are
@@
 
-\begin{verbatim}
+\begin{small}
+\begin{Verbatim}
 data Z
@@
   (:.) :: Shape tail -> Int -> Shape (tail :. Int)
-\end{verbatim}
+\end{Verbatim}
+\end{small}
 
@@
 
-\begin{verbatim}
+\begin{small}
+\begin{Verbatim}
 data Pull sh a = Pull (Shape sh -> a) (Shape sh)
@@
  Pull (\(sh' :. i) -> ixf (sh' :. (l - i - 1))) sh
-\end{verbatim}
+\end{Verbatim}
+\end{small}
 
@@
 
-Pull arrays on the other hand were conveived in the context of a
+Pull arrays on the other hand were conceived in the context of a
 typed functional language. The notation is completely inherited from
@@
 The correspondence between MoA and Pull arrays is most clear when
-considering Denotaional Normal Form (DNF), \(\psi\)-reductions and
+considering Denotational Normal Form (DNF), \(\psi\)-reductions and
 fusion. Both formalisms define operations on arrays by how to index
@@
 used in MoA and Pull arrays. The intention of this figure is to work
-as a rosetta stone, helping to translate from one formalism to the
+as a Rosetta stone, helping to translate from one formalism to the
 other.

### 2014-04-11 paper.tex
An abstract

@@
 \begin{abstract}
-This is the text of the abstract.
+The aim of this paper is to rectify some of the fragmentation
+high-performance array programming community. We look at two
+formalisms for functional, high-performance array computations:
+Mathematics of Arrays and Pull arrays.  On the surface they appear
+rather different but at their core they share many principles and
+properties.  In this paper we spell out the striking similarities
+between these two formalisms. This paper is meant as a guide for
+researchers knowledgeable in one of the formalisms who would like to
+learn about the other.
+
 \end{abstract}

### 2014-04-11 paper.tex
updated the abstract
What do you think?

--- 2014-04-11 paper.tex
@@
 \begin{abstract}
-This is the text of the abstract.
+The aim of this paper is to rectify some of the fragmentation
+within the high-performance array programming community. We look at two
+formalisms for functional, high-performance array computations:
+Mathematics of Arrays and Pull arrays, two seemingly different formalisms
+that share many principles and properties.   This paper is meant to
+identify the striking similarities between the two  and to act as a guide  to enable research communications.
+
 \end{abstract}

### 2014-04-12 paper.tex
one type fixed
I also edited the footnote. Everything looks good.
Goodnight.

@@
 removed. But sometimes it is important to keep an array in memory
-\footnote{Example: Indexing a huge transposed array that has a few rows and many columns. Non-materialization
-may cause many cache misses and page faults.}, 
+\footnote{Example: Indexing a column of a huge transposed array that has a few rows and many columns. In this case, non-materialization
+may have long strides causing many cache misses and page faults.}, 
 not least to produce the final result. For this reason, Pull arrays
@@
 
-Pull arrays are always kept on the form of an indexing function and a
+Pull arrays are always kept in the form of an indexing function and a
 length. This representation is preserved for all operations. In MoA,

### 2014-04-14 paper.tex
I found a few more typos
Nothing important but I found two more typos when I read it again.

@@
 is challenging in a typed language such as Haskell because it leads to
-a form of infinite recursion and while there are a number of way to
+a form of infinite recursion and while there are a number of ways to
 break the recursion we have yet to find a way that does so and retains

### 2014-04-14 intromoa.tex
I found a few more typos
Nothing important but I found two more typos when I read it again.
Fixing a dyslexic event
In my addition derivation I corrected places i had
4 5 instead of 5 4. Some were correct but I just found these.

@@
 
-Note
-also that all operations are over the 0th axis. The operator $\Omega$~\cite{mul88} extends operations over all other dimensions.
-
 \subsection*{Higher Order Operations}
@@
 performed over their 0th dimensions.  We introduce the higher order binary 
-operation $\Omega$, which is defined when its left argument is a unary or binary
+operation $\Omega$~\cite{mul88}  which extends operations over all dimensions.
+$\Omega$ is defined when its left argument is a unary or binary
 operation and its right argument is a vector describing the dimension upon 
@@
 &=& 2 \cat <\;5\;4> \nonumber \\
-&=& <2\;4\;5>
+&=& <2\;5\;4>
 \end{eqnarray}
 Now that we have the {\em shape}, we can {Psi-Reduce}.
-\[ \forall \;\;\; i,j,k \;\;\; \ni \;\;\; 0 \;\;\;\leq i,j,k < \;\;\;<2\;4\;5> \]
+\[ \forall \;\;\; i,j,k \;\;\; \ni \;\;\; 0 \;\;\;\leq i,j,k < \;\;\;<2\;5\;4> \]
 \[<i\;j\;k> \psi X \]
@@
 i.e. the bounds for $j$ and $k$  above.
-Continuing with the PCT to create the {\em generic} ONF on p=2 processors, we use the index of the primary axis, i, $\ni\;\;0 \leq  i < p$, since $\rho X = <2\;4\;5>$.
+Continuing with the PCT to create the {\em generic} ONF on p=2 processors, we use the index of the primary axis, i, $\ni\;\;0 \leq  i < p$, since $\rho X = <2\;5\;4>$.
 Using $\gamma_{row}$ for row major ordering, Y is transformed into:
@@
 a[ (( \gamma(<2-i >; <1 \take (\rho A>))) \times \pi ( 1 \drop (\rho A))   )  + \iota \pi 1 \drop \rho A  ] \nonumber \\
-\times a[ (( \gamma(<1-i >; <1 \take (\rho A>))) \times \pi ( 1 \drop (\rho A))   )  + \iota \pi 1 \drop \rho A  ] 
+\times a[ (( \gamma(<1-i >; <1 \take (\rho A>))) \times \pi ( 1 \drop (\rho A))   )  + \iota \pi 1 \drop \rho A  ] \nonumber 
 \end{eqnarray}
@@
 \times a[ (( \gamma(<1-i >; <3>))) \times \pi ( <5\;4>)   )  + \iota \pi <5\;4>  ]\nonumber \\
-a[( (2-i) \times  20)  + \iota 20 ] \nonumber \\
+=\;\;a[( (2-i) \times  20)  + \iota 20 ] \nonumber \\
 \times a[( (i-1) \times 20 )  + \iota 20 ]
